Quantifying the Impact \\ of Performance Improvements and Cost Reductions \\ from 20 years of Light-Emitting Diode Manufacturing

### 2021-02-11 SPIE/article/main.tex
more corrections

    The early development history of light-emitting diodes is well documented \cite{Steigerwald2002}\cite{Nakamura2013}\cite{feezell2018invention}. Major works dealing with device efficiency improvements have documented overall progress \cite{azevedo2009transition}\cite{tsao2010solid}\cite{pattison2017solid}. Despite the excellent overview these publications provide, they did not systematically investigate the sources and impact of related technology improvements. Investigations regarding cost reductions are rarely reported in scientific literature. The focus is instead on the energy economics implications of decreased electricity consumption \cite{tsao2010world}\cite{Tsao2010SSL}.

\clearpage
\section{METHODS}
\label{sec:methods}

        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed. It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures. We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture. An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}. The model was constructed for the years 2003, 2012 and 2020. It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location. Process specific step parameters were derived from scientific literature, company publications, archived product catalogs and patent literature. Where necessary, this data was augmented using information gathered during semi-structured interviews with experts from industry and academia. Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by semi-structured interviews with experts from industry and academia. For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}. Our model includes both the wafer treatment process as well as the packaging process. While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach. The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.

        \begin{figure} [ht]
            \begin{center}
                \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
            \end{center}
            \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading). Note that dimensions are not to scale and smaller features are greatly exaggerated for visibility. Years correspond to earliest identified patent priority date. Dashed boxes indicate chip designs not brought to large scale production. Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
            \label{fig:chip_arch}
        \end{figure}

        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al. in a cost model for photovoltaic modules \cite{kavlak2018evaluating}. It is based on the logarithmic derivative of the total differential of the cost function. The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can thus be approximated. For the detailed derivation, we refer to the supplementary material of the original publication.

    \subsection{Data Sources and Performance Calculations}+

        XXX mention patent analysis
        XXX mention interviews
        XXX mention company publications
        XXX move any relevant sentences from cost model subsection here

        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in devices, we gathered data on overall device performance and identified the associated device architecture, manufacturing processes and types of down conversion phosphors used. We systematically identified the architecture and manufacturing process improvements related to each physical loss channel, gathered associated efficiency data from literature or computed the respective values from raw data.

@@

@@

    Once technological breakthroughs and process improvements had been identified, the next step was to quantify their effect. The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels. They must further have the ability to directly describe the effect of single technological breakthroughs, device architectures and manufacturing process improvements on device performance.

    Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, to demonstrate the limitations of these metrics. It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.

    \begin{figure} [ht]
        \begin{center}
            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
        \end{center}
        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}. Note that flux per single chips has not increased as starkly as the flux for multi-chip packages. Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals. Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate. Sources: Data from Weinold et al. \cite{weinold2020technology}}
        \label{fig:haitz}
    \end{figure}

    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency. This metric is defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$. Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements. Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.

@@

     Results for performance improvements are presented in Figure \ref{fig:efficiency}. Overall efficiency in best performing devices improved from $\eta_L=5.8\%$ in 2002 to $\eta_L = 38.7\%$ in 2020. As previous studies have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}. We note, however, that those loss channels with a fixed physical limit become significantly more dominant in 2016 and 2020. For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant. Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels. Notably, sub-efficiencies for the most current devices are only $\sim10\%$ below the physical limit. The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for improvement.

@@

     XXXXXXXXXXXXXXXX

@@

\clearpage
\begin{figure} [ht]
    \begin{center}
        \includegraphics[width=0.85\textwidth]{SPIE/article/breakthroughs_efficiency.pdf}
    \end{center}
    \caption{Impact of technology breakthroughs and manufacturing process improvements on historical developments in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA. The overall lamp efficiency $\eta_L$ is displayed as the rightmost column. This figure takes as inputs the state-of-the-art sub-efficiencies discussed in section \ref{sec:methods}. Horizontal colored bars give state-of-the-art sub-efficiencies for five years: \textcolor{blue}{1997}, \textcolor{teal}{2002}, \textcolor{orange}{2010}, \textcolor{magenta}{2016} and \textcolor{red}{2020}. Colored annotation "N/A" indicates that the sub-efficiency of the corresponding year cannot be computed for the following reasons: V$_\text{f}$E, Droop: depend on current, which was below 350mA at the time; CE, SE: warm white spectrum LEDs not available at the time. Physical limits are indicated by black horizontal bars. The possible range for the physical limit of V$_\text{f}$E exceeds 100\% and depends on electrical device parameters, which are discussed in \cite{david2016electrical}. The range of this limit is indicated by an upward pointing black arrow. Efficiency acronyms are listed in table \ref{tab:eff}. Data for historical improvements of sub efficiencies from Weinold et al. \cite{weinold2020technology}}
    \label{fig:efficiency}
\end{figure}

>>>>

\clearpage
\section{Methods and Data}
\label{sec:methods}

    To describe the effect of technological breakthroughs and manufacturing process improvements, a set of metrics had to be identified. The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels. They must further have the ability to directly describe the effect of single technological breakthroughs, evolution in device architecture and manufacturing process improvements on device performance.

    Figure \ref{fig:haitz} shows an updated and expanded overview of the flux per device and the cost per flux of highest performing light-emitting diodes, both at the chip and package level, to demonstrate the limitations of these metrics. It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.

    \begin{figure} [ht]
        \begin{center}
            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
        \end{center}
        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}. Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals. Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate.}
        \label{fig:haitz}
    \end{figure}

    \clearpage
    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency. This metric is defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$. Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements. Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.

@@

    \subsection{Data Sources and Performance Calculations}
    \label{subsec:data}

        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in light-emitting diode manufacturing, we gathered data from a variety of sources. The aim was to collect historical data on overall device performance and historical development in sub-efficiencies. Where possible, we gathered  efficiency data from literature or computed the respective values from raw data. We further had to identify the associated device architectures, manufacturing processes and types of down conversion phosphors used. Data was extracted from patent literature, scientific publications, industry reports and technology forecasts. Additional information and data was provided by experts from both the light-emitting diode manufacturing industry and the manufacturing equipment industry during semi-structured interviews.

@@

        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed. It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures. We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture. An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}. The earliest commercial warm-white phosphor converted blue light emitting diodes were introduced in 2003, so the model was constructed for the years 2003, 2012 and 2020. It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location. Details of the manufacturing processes and process-specific step parameters were derived from the sources mentiond in section \ref{subsec:data}. For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}. Our model includes both the wafer treatment process as well as the packaging process. While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach. The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.

        \begin{figure} [ht]
            \begin{center}
                \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
            \end{center}
            \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading). Note that dimensions are not to scale. Years correspond to earliest identified patent priority date. Dashed boxes indicate chip designs not brought to large scale production. Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
            \label{fig:chip_arch}
        \end{figure}

        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al. in a cost model for photovoltaic modules \cite{kavlak2018evaluating}. It is based on the logarithmic derivative of the total differential of the cost function.  For the detailed derivation, we refer to the supplementary material of the original publication.

@@

     Results for performance improvements are presented in Figure \ref{fig:efficiency}. Overall efficiency in best performing devices improved from $\eta_L=5.8\%$ in 2002 to $\eta_L = 38.7\%$ in 2020. As previous studies have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}. Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels. We note, that those loss channels with a fixed physical limit below 100\% have become significantly more dominant in 2016 and 2020. For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant. Notably, sub-efficiencies for the most current devices are only $\sim10\%$ below the physical limit. The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for improvement.

@@

     XXX video text

    \begin{figure} [ht]
        \begin{center}
            \includegraphics[width=0.85\textwidth]{SPIE/article/breakthroughs_efficiency.pdf}
        \end{center}
        \caption{Impact of technology breakthroughs and manufacturing process improvements on historical developments in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA. The overall lamp efficiency $\eta_L$ is displayed as the rightmost column. This figure takes as inputs the state-of-the-art sub-efficiencies discussed in section \ref{sec:methods}. Horizontal colored bars give state-of-the-art sub-efficiencies for five years: \textcolor{blue}{1997}, \textcolor{teal}{2002}, \textcolor{orange}{2010}, \textcolor{magenta}{2016} and \textcolor{red}{2020}. Colored annotation "N/A" indicates that the sub-efficiency of the corresponding year cannot be computed for the following reasons: V$_\text{f}$E, Droop: depend on current, which was below 350mA at the time; CE, SE: warm white spectrum LEDs not available at the time. Physical limits are indicated by black horizontal bars. The possible range for the physical limit of V$_\text{f}$E exceeds 100\% and depends on electrical device parameters, which are discussed in \cite{david2016electrical}. The range of this limit is indicated by an upward pointing black arrow. Efficiency acronyms are listed in table \ref{tab:eff}. Data for historical improvements of sub efficiencies from Weinold et al. \cite{weinold2020technology}}
        \label{fig:efficiency}
    \end{figure}

@@
